Automated curriculum learning for Multi-Domain NMT by using loss-driven reward

### 2021-05-06 2021/emnlp/auto_cl_mdmt.tex
adding new version of the paper, adding scores, adding figures, adding new references

\documentclass[11pt,a4paper]{article}
\usepackage[hyperref]{acl2021}
\usepackage{times}
\usepackage{latexsym}
\renewcommand{\UrlFont}{\ttfamily\small}

@@

@@

\begin{document}
\maketitle
\setlength{\abovedisplayskip}{2pt}
\setlength{\belowdisplayskip}{2pt}
\begin{abstract}
  When building machine translation systems, one often needs to make the best out of heterogeneous sets of parallel data in training to achieve the best test performance in one or several domain(s) of interest. This multi-source / multi-domain adaptation problem is often approached with instance selection or reweighting strategies, most of which pre-suppose an ex-ante assessment of the relevance of training instances. In this paper, we propose a novel dynamic data selection algorithm which re-evaluates the data utility and adapts the data selection policy along the training of the NMT model. We explore the ability of our method to serve as a generic framework for the previously mentioned situations.
\end{abstract}

\section{Introduction}\label{sec:intro}
A typical setting in machine translation (MT) is to collect the largest possible collection of parallel data for the chosen language pair, with the intent to achieve optimal performance for a task of interest. In such situations, the training data distribution is opportunistic, while the test data distribution is chosen and fixed; a key component in training is then to mitigate the detrimental effects of a possible mismatch between these distributions. Single-source and multi-source domain adaptation (DA) is a well-studied instance of this setting (see \citep{Chu2017comparison} for a review) , and so is multi-domain (MD) learning \cite{Chu18multilingual,Zeng18multidomain,Jiang19multidomain,Pham21revisiting}. A related situation is multilingual machine translation \cite{Firat16multiway,Ha16towards,Johnson17google,Arivazhagan19massively}\fyTodo{Add more recent work}, where the heterogeneity of training data not only corresponds to variations in topic, genre or register, but also in language.

This problem is often approached by \emph{static} instance selection or re-weighting strategies, where the available training data is used in proportion to its relevance for the chosen test conditions. Finding the optimal balance of training data is however a challenging task due, for instance, to similarity between domains / languages, but also due to regularization effects of out-of-domain data; it may also be suboptimal, as some target domains or languages might be easier to train than others. Several recent proposals \cite{Wang17instance,Zhang19curriculum,Kumar19reinforcement,Wang20learning-multi,alex17automated} have explored ways to instead consider \emph{dynamic} data selection and sampling strategies. In this paper, we propose a nouvel method of building an automated curriculum for training Multi-Domain NMT model, that is inspired by the work of \cite{alex17automated}. We evaluate our method in various single-domain and multi-domain adaptation settings. We compare our method with several contrasts including the work of \cite{Zhang19curriculum} and \cite{Wang20balancing}. From now on, we name our method MD-ACL

Based on experimental results obtained on a diverse set of domains, our main conclusions are that (a) using MD-ACL often yields overall performance that are as good as the standard fine-tuning strategy for domain adaptation; (b) MD-ACL can effectively handle a variety of test target distributions without any meta-parameter search.
%(c) our extension of DDS is able to boost its performance when used in conjunction to the adapter model of \citet{Bapna19simple}.
\section{Learning with multiple data sources} \label{sec:mdmt}

@@

The main challenge is then to make the best of this heterogeneous data, with the aim to achieve the optimal performance for the intended test conditions. These might correspond to data from just one of the training domains, as in standard supervised (multi-source) domain adaptation; a more difficult case is when the test data is from one domain unseen in training (unsupervised domain adaptation); in multi-domain adaptation finally the test distribution is itself a mixture of domains, some of which may also be observed in training.  Without loss of generality, one may then assume that the test distribution takes the form $\mathcal{D}^{t}(x) = \sum_d \lambda^{t}(d) \mathcal{D}_d(x)$ - with only one non-null  component in the case of fine-tuning.
These situations are illustrated in Figure~\ref{fig:mdmt-lambdas}.
\begin{figure}[h]
  \centering
  \includegraphics[width=0.48\textwidth]{mdmt-lambdas}
  \caption{Training and testing with distribution mismatch. We consider just three domains, and represent vectors of mixture weights $\vlambda^{s}$ and $\vlambda^{t}$ in the 3-dimensional simplex. Training with weights in (a) and testing with weights in (c) is supervised multi-source domain adaptation to domain~2 ($d_2$), while (b)-(c) is the unsupervised version, with no training data from $d_2$; training with weights in (a) and testing with weights in (d) is multi-domain learning, also illustrated with configurations (a)-(e) (training domain $d_1$ is not seen in test), and (b)-(d)  (test domain $d_2$ is unseen in training).}\label{fig:mdmt-lambdas}
\end{figure}

These situations have been amply documented from a theoretical perspective (eg.\ \cite{Mansour09multiple,Mansour09domainadaptation,Hoffman18algorithms}). A general recommandation in the DA setting  is to adjust the sampling distribution used to optimize the system so as to compensate for the mismatch between $\mathcal{D}^s(x)$ and $\mathcal{D}^t(x)$. This can be approximated by reweighting instances, or more conveniently domains, which are selected during training with a probability $\lambda^{l}(d)$, with $\lambda^{l}(d) \neq \lambda^{s}(d)$.

A more practical standard approach to supervised DA is \emph{fine-tuning} \cite{Luong15stanford,Freitag16fast}, where $\vlambda^{l}$ is allowed to vary during the course of learning. With our notations, this approach amounts to first learning an initial parameter value with all the data ($\forall d, \lambda^{l}(d) = \lambda^{s}(d)$), then to continue training with only batches from the test domain $d_t$ ($\lambda^{l}(d) = \indic{d = d_t}$) with $\indic{A}$ the indicator function for predicate $A$. Note that this strategy is potentially suboptimal, as some out-of-domain samples may contribute to the final performance due to eg.\ domain overlap. Optimizing the learning distribution in multi-domain settings is even more challenging, as the learner has to best take advantage of potential domains overlaps, and also of the fact that some domains might be easier to learn than others.\fyTodo{How to measure this?} \revision{Answer: The graphic of \system{Mixed-0} shows that with same amount of data(batches) the convergences of the domains vary significantly}

Differentiable data selection (DDS), proposed by \citet{Wang20optimizing} and later applied to multilingual machine translation by \citet{Wang20balancing}, introduces an additional degree of freedom and allows to iteratively learn an optimal value for $\lambda^{l}$, as we now explain.

@@

\section{Differential Data Selection} \label{sec:dds}
\subsection{Basic principles}
Assuming training data from $n_d$ domains $d_1 \dots d_{n_d}$, we denote the size of the training corpus from domain $d$ as  $N^{s}_d$, and $N^{s} = \sum_d N^{s}_d$ is the total number of training samples. We use $\widehat{\mathcal{D}^l_d}$ and $\widehat{\mathcal{D}^t _d}$ to denote the empirical train and test distributions for domain $d$, and \revision{$\widehat{\mathcal{D}^{u}}(x;\lambda^{u}) = \sum_{d} \lambda^{u}(d) \widehat{\mathcal{D}^{u}_d}(x)$ for $u\in\{l,t\}$} \fyTodo{$\widehat{\mathcal{D}}(x;\lambda) = \sum_{d} \lambda(d) \widehat{\mathcal{D}_d}(x)$}. In our setting,  $\vlambda^t$, and hence $\widehat{\mathcal{D}^t}(x;\vlambda^t)$ are fixed and predefined, approximated with an equivalent number of development corpora. Training optimizes the parameters $\vtheta$ of a neural translation model.

Sampling training domains using a distribution $\lambda^{l}$ results in a parameter estimate $\theta(\lambda^{l})$, delivering performance that can be estimated on the development set using $\lambda^{t}$. \citet{Wang20optimizing,Wang20balancing} then formulate the problem of finding an optimal sampling strategy $\lambda^{l}_{*}$ maximizing performance on the development data as a bi-level optimization problem that can be approached by coordinate gradient descent alternating between $\lambda^{l}$ and $\vtheta$. \citet{Wang20balancing} limit the search space of $\vlambda^{l}$ to a parameterized differentiable function $\vlambda^{l}(\vpsi)$. Given $\vpsi_t$ at time step $t$, $\vtheta_t$ is updated via usual gradient descent as follows.
\begin{align*}
\vtheta_{t+1} &= \vtheta_t - \mathbf{lr}_{nmt} * \frac{1}{N} \displaystyle{\mathop{\sum}_{i=1}^N} \frac{\partial l(\vtheta_t, x_i,y_i)}{\partial \vtheta} \\
x_i,y_i &\sim \widehat{\mathcal{D}^l}(x;\vlambda^l(\psi_t)) = \displaystyle{\mathop{\sum}_{d=1}^K} \lambda^l(d;\psi_t) * \widehat{D^l_d}(x) \\
\end{align*}
Where $l(\theta_t,x,y)$ is the loss on an example.

Given $\vtheta_t$, $\psi_t$ is updated as follows.
%via the Reinforce algorithm \cite{Williams92simple}:\fyTodo{Rewrite this correctly with domain average reward}
\begin{align*}
\psi_{t+1} &= \psi_t + \mathbf{lr}_{data} * \displaystyle{\mathop{\sum}_{d=1}^K} R(d) * \frac{\partial log(\lambda^l(d;\psi_t))}{\partial \psi} \\
  R(d) &= \langle \frac{\partial J^l_d(\theta_t)}{\partial \theta}, \frac{\partial J^t(\theta', \lambda^t)}{\partial \theta} \rangle \\
\end{align*}
Where
\begingroup
\allowdisplaybreaks
\begin{align*}
\theta' &= \theta_t - \mathbf{lr}_{nmt} * \displaystyle{\mathop{\sum}_{i=1,x^i_d,y^i_d \in \widehat{D^l_d}}^N} \frac{\partial l(\theta_t,x^i_d,y^i_d)}{\partial \theta} \\
  J^l_d(\theta_t) &= \displaystyle{\mathop{\sum}_{i=1,x^i_d,y^i_d \in \widehat{D^l_d}}^N} l(\theta_t,x^i_d,y^i_d) \\
  J^t(\theta',\lambda_t) &= \displaystyle{\mathop{\sum}_{d}}\lambda^t(d)\displaystyle{\mathop{\sum}_{x^t_d,y^t_d \in \widehat{D^t_d}}} l(\theta',x^t_d,y^t_d)\\
\end{align*}
\endgroup
$\mathbf{lr}_{data}$ is the learning rate of the parameter of the sampler. $\mathbf{lr}_{nmt}$ is the learning rate of the parameter of the NMT system. \fyTodo{Explain lr, may be more}\fyTodo{More refs to the algorithm }
% To apply this method to multi-domain training (\system{Multi-DDS}), we limit the search family of training distribution $D_{trn}$ to the family of mixture of K in-domain training distribution such as $\sum_d \lambda^{s}_{d} \mathcal{D}_d(x,y)$. We also parameterize $\lambda^{s}_{d}$ using Softmax function so that $\lambda^{s}_{d} = \text{Softmax}(\psi)_d$ where $\psi \in \mathbb{R}^K$.
\subsection{DDS for (multi) domain adaptation}

This setting is pretty general and can in principle accomodate the variety of situations mentioned  above, and many more: basic domain adaption, multi-domain adaptation with various target distributions, possibly including domains unseen in training. In our experiments we would like to better assess the actual potential of DDS is these settings and seek to experimentally answer the following questions:
\begin{itemize}
\item is DDS a viable alternative to conventional fine-tuning? In particular does it enable to better take advantage of relevant data from other domains?
\item is DDS also a viable option in multidomain adaptation scenarios?
\item does DDS also enable to perform unsupervised (multi-) domain adaptation?\fyTodo{TBContinued}
\end{itemize}

@@

\section{Experimental settings} \label{sec:exp}

\subsection{Data and metrics \label{ssec:corpora}}
We experiment with translation from English into French and use texts initially originating from 6~domains, corresponding to the following data sources: the UFAL Medical corpus V1.0 (\domain{med})\footnote{\url{https://ufal.mff.cuni.cz/ufal_medical_corpus}.\revision{We only use the in-domain (medical) subcorpora: PATR, EMEA, CESTA, ECDC.}}; the European Central Bank corpus (\domain{bank}) \cite{Tiedemann12parallel}; the JRC-Acquis Communautaire corpus (\domain{law}) \cite{Steinberger06acquis}; documentations for KDE, Ubuntu, GNOME and PHP from the Opus collection \cite{Tiedemann09news}, merged in a \domain{it}-domain; TedTalks (\domain{talk}) \cite{Cettolo12wit}, and the Koran (\domain{rel}). Complementary experiments also use v12 of the News Commentary corpus (\domain{news}). Most corpora are available from the Opus website.\footnote{\url{http://opus.nlpl.eu}} These corpora were deduplicated and tokenized with in-house tools; statistics are in Table~\ref{tab:Corpora}. To reduce the number of types and build open-vocabulary systems, we use Byte-Pair Encoding \cite{Sennrich16BPE} with 30,000 merge operations on a corpus containing all sentences in both languages.\fyDone{Add \# number of tokens, also specificity ?}%

@@

%for contrast experiments, we also use supplementary test sets from three other domains: the official Khresmoi testset \cite{Khresmoi17test}, which is close to EMEA, News test 2014 \cite{Bojar14findings}, and IWSLT 2010 (Talk track) \cite{Paul10overview}. This enables us to evaluate the loss in performance when the test set is from a domain not seen in training.
% The model is also required to achieve comparable performance to generic model. To do so, we use newstest 2009 and IWSLT 2010 whose contain does not particularly belong to any domain.

@@

\fyTodo{Keep this ?}
We measure the distance between domains using the $\mathcal{H}$-Divergence \cite{Ben-David09atheory}, which relates domain similarity to the test error of a domain discriminator: the larger the error, the closer the domains.
%Formally, given test sets of size $m$ for domains $A$ and $B$, and $h(x)$ a trained domain predictor, $\mathcal{H}(A,B)$ is computed %as:
%$$
%\mathcal{H}(A,B) = 2(1 - [\frac{1}{m} \sum_{x:h(x) = B} \mathbb{I}( x \in A) + \frac{1}{m} \sum_{x: h(x) = A} \mathbb{I}(x \in B)]),
% $$
% where $\mathbb{I}$ is the indicator function.
Our discriminator is a SVM independently trained for each pair of domains, with sentence representations derived via mean pooling from the source side representation of the generic Transformer model. We used the scikit-learn\footnote{\url{https://scikit-learn.org}} implementation with default values.\fyDone{Inform the classifier details}\fyDone{Insert tableau} Results in Table~\ref{tab:domaindist} show that all domains are well separated from all others, with \domain{rel} being the furthest apart, while \domain{talk} is slightly more central.

\begin{table}\centering
  \begin{tabular}{|l*{5}{|r}|} 
  \cline{2-6}
  \multicolumn{1}{c|}{} & \domain{law} & \domain{bank} & \domain{talk} & \domain{IT} & \domain{rel} \\ \hline
    \domain{med} &1.93 &1.97 &1.9 &1.93 &1.97 \\
    \domain{law}   && 1.94 & 1.97 &1.93 & 1.99 \\
    \domain{bank} &&&1.98 &1.94 &1.99 \\
    \domain{talk}   &&&&1.92 &1.93 \\
     \domain{IT}     &&&&& 1.99 \\ \hline
  \end{tabular}
  \caption{The $\mathcal{H}$-divergence between domains}
  \label{tab:domaindist}
\end{table}

@@

\begin{itemize}
\itemsep0em 
\item Generic models trained with various predefined mixtures of the training data taking the form:
\begin{align} \label{mixture:trn}
\lambda_{\alpha}(d) = \frac{q_d^{\alpha}}{\displaystyle{\mathop{\sum}_{d=1}^{n_d}q_d^{\alpha}}} &&
q_d = \frac{\mid N^{s}_d \mid}{\displaystyle{N^{s}}} % \mathop{\sum}_{i=1}^K\mid D_i \mid}}
\end{align} 
with $\alpha \in [0,0.25,0.5,0.75,1.0]$. These systems are denoted \system{Mixed-$\alpha$} below. \system{Mixed-$0$} uses a uniform domain distribution, \system{Mixed-$1.0$} simple uses the observed domain distribution.
%on a concatenation of all corpora (\texttt{Mixed}). We develop two versions\footnote{In fact three: to enable a fair comparison with WDCMT, a RNN-based variant is also trained and evaluated. \revision{This system appears as \system{Mixed-Nat-RNN} in Table~\ref{tab:performance}}.} of this system, one where the domain unbalance reflects the distribution of our training data \revision{given in Table~\ref{tab:Corpora}} (\system{Mixed-Nat}) and one where all domains are equally represented in training (\system{Mixed-Bal}). The former is the best option when the train mixture $\mathcal{D}^s$ is also expected in testing; the latter should be used when the test distribution is uniform across domains. Accordingly, we report two aggregate scores: a weighted average reflecting the training distribution, and an unweighted average, meaning that test domains are equally important.
\item fine-tuned models \cite{Luong15stanford,Freitag16fast}, based on the \system{Mixed-$1.0$} system, further trained on each domain for at most 50~000 iterations, with early stopping when the dev BLEU stops increasing. The full fine-tuning (\system{FT-Full}) procedure may update all the parameters of the initial generic model, resulting in six systems each adapted for one domain, with no parameter sharing across domains.

\item two multi-domain versions of the approach of \newcite{Bapna19simple}, denoted \system{FT-Res} and \system{MDL-Res}, where a domain-specific adaptation module is added to all the Transformer layers; within each layer, residual connections enable to short-cut this adapter. The former variant corresponds to the original proposal of \citet{Bapna19simple} (see also \cite{Sharaf20metalearning}). It fine-tunes the adapter modules of a \system{Mixed-$1.0$} system independently for each domain, keeping all the other parameters frozen. The latter uses the same architecture, but a different training procedure and learns all parameters jointly from scratch with the prefixed mixtures of training data \ref{mixture:trn}.\fyTodo{Keep this ?}

\item Our comparison of multi-domain systems includes baselines with fixed data mixtures corresponding to $\vlambda_0$ and $\vlambda_1$, and our own reimplementations of recent proposals from the literature:\footnote{\revision{Further implementation details are in supplementary material.}}\fyTodo{Choose a reasonnable MD baseline }

\end{itemize}

@@

\subsection{DDS-based systems } \label{ssec:dds-sys}

The architecture of DDS based-systems is identical to that of our baselines, and only differ in the training regime. Their behavior only depends on (a) the initial domain distribution at the start of training $\vlambda^{l}_{t=0}$, and (b) the targeted (dev/test) distribution $\vlambda^{t}$. We will thus report these systems as \system{DDS($\vlambda^{l}_{t=0}$, $\vlambda^{t}$)}. The domain distribution attained at the end of training will be denoted  $\vlambda^{l}_{*}$.

@@

\subsection{DDS for single-domain and multi-domain adapation}

\subsubsection{Fine-tuning experiments}

\fyTodo{Fix the table}
In Table ~\ref{tab:basic-dds}, we compare the results of fine-tuning a system initially trained with a natural mixture, which is proportional to the sizes of the domains, of domains with the result of DDS learning from scratch with a initial uniform distribution.

>>>>

\documentclass[11pt,a4paper]{article}
\usepackage[hyperref]{acl2021}
\usepackage{times}
\usepackage{latexsym}
%\renewcommand{\UrlFont}{\ttfamily\small}

@@

\let\oldbibitem\bibitem
\def\bibitem{\vfill\oldbibitem}

@@

\begin{document}
\maketitle
\setlength{\abovedisplayskip}{2pt}
\setlength{\belowdisplayskip}{2pt}
\begin{abstract}
  When building machine translation systems, one often needs to make the best out of heterogeneous sets of parallel data in training to achieve the best test performance in one or several domain(s) of interest. This multi-source / multi-domain adaptation problem is often approached with instance selection or reweighting strategies, most of which pre-suppose an ex-ante assessment of the relevance of training instances. In this paper, we propose a novel dynamic data selection algorithm that re-evaluates the data utility and adapts the data selection policy along with the training of the NMT model. We explore the ability of our method to serve as a generic framework for the previously mentioned situations.
\end{abstract}

\section{Introduction}\label{sec:intro}
A typical setting in machine translation (MT) is to collect the largest possible collection of parallel data for the chosen language pair, with the intent to achieve optimal performance for a task of interest. In such situations, the training data distribution is opportunistic, while the test data distribution is chosen and fixed; a key component in training is then to mitigate the detrimental effects of a possible mismatch between these distributions. Single-source and multi-source domain adaptation (DA) is a well-studied instance of this setting (see \citet{Chu2017comparison} for a review), and so is multi-domain (MD) learning \citep{Chu18multilingual,Zeng18multidomain,Jiang19multidomain,Pham21revisiting}. A related situation is multilingual machine translation \citep{Firat16multiway,Ha16towards,Johnson17google,Arivazhagan19massively}\fyTodo{Add more recent work}, where the heterogeneity of training data not only corresponds to variations in topic, genre or register, but also in language.

This problem is often approached by \emph{static} instance selection or re-weighting strategies, where the available training data is used in proportion to its relevance for the chosen test conditions \citep{Moore10selection,Axelrod11domain}. Finding the optimal balance of training data is however a challenging task due, for instance, to the similarity between domains/languages, but also due to regularization effects of out-of-domain data; it may also be suboptimal, as some target domains or languages might be easier to train than others. For example, improving the performance of the MT system in one domain will often hurt that of another \citep{Vanderwees17dynamic, Britz17mixing}, and improving model generalization across all domains \citep{koehn18findings} may not promise optimization of a particular domain.

Several recent proposals \citep{platanios19competence,Zhang19curriculum,Kumar19reinforcement,Wang20learning-multi,Graves17automated,Wang20balancing,Wang20learning-multi,Vanderwees17dynamic} have explored ways to instead consider \emph{dynamic} data selection and sampling strategies that surpass static strategies. Furthermore \citet{Zhang19curriculum,Vanderwees17dynamic} construct a fixed curriculum while \citet{platanios19competence,Kumar19reinforcement,Graves17automated,Wang20balancing,Wang20learning-multi} build curricula automatically adapted to the training.

In this paper, we propose a novel method of building an automated curriculum for training the Multi-Domain NMT model, that is inspired by the work of \citet{Graves17automated}. We evaluate our method in various single-domain and multi-domain adaptation settings. We compare our method with several contrasts including the work of \citet{Zhang19curriculum}, that we name CL, and \citet{Wang20balancing}, that we name DDS. From now on, we name our method MDAC which is brief for Multi-Domain Automated Curriculum. Based on experimental results obtained on a diverse set of domains, our main conclusions are that (a) using MDAC often yields overall performance that are as good as the standard fine-tuning strategy for domain adaptation; (b) MDAC can effectively handle a variety of test target distributions without any meta-parameter search.

\section{Learning with multiple data sources} \label{sec:mdmt}

@@

The main challenge is then to make the best of this heterogeneous data, with the aim to achieve the optimal performance for the intended test conditions. These might correspond to data from just one of the training domains, as in standard supervised (multi-source) domain adaptation; a more difficult case is when the test data is from one domain unseen in training (unsupervised domain adaptation); in multi-domain adaptation finally, the test distribution is itself a mixture of domains, some of which may also be observed in training.  Without loss of generality, one may then assume that the test distribution takes the form $\mathcal{D}^{t}(x) = \sum_d \lambda^{t}(d) \mathcal{D}_d(x)$ - with only one non-null  component in the case of fine-tuning.
These situations are illustrated in Figure~\ref{fig:mdmt-lambdas}.
\begin{figure}[h]
  \centering
  \includegraphics[width=0.48\textwidth]{mdmt-lambdas}
  \caption{Training and testing with distribution mismatch. We consider just three domains, and represent vectors of mixture weights $\vlambda^{s}$ and $\vlambda^{t}$ in the 3-dimensional simplex. Training with weights in (a) and testing with weights in (c) is supervised multi-source domain adaptation to domain~2 ($d_2$), while (b)-(c) is the unsupervised version, with no training data from $d_2$; training with weights in (a) and testing with weights in (d) is multi-domain learning, also illustrated with configurations (a)-(e) (training domain $d_1$ is not seen in test), and (b)-(d)  (test domain $d_2$ is unseen in training).}\label{fig:mdmt-lambdas}
\end{figure}

These situations have been amply documented from a theoretical perspective (eg.\ \cite{Mansour09multiple,Mansour09domainadaptation,Hoffman18algorithms}). A general recommendation in the DA setting is to adjust the sampling distribution used to optimize the system so as to compensate for the mismatch between $\mathcal{D}^s(x)$ and $\mathcal{D}^t(x)$. This can be approximated by reweighting instances, or more conveniently domains, which are selected during training with a probability $\lambda^{l}(d)$, with $\lambda^{l}(d) \neq \lambda^{s}(d)$.

A more practical standard approach to supervised DA is \emph{fine-tuning} \cite{Luong15stanford,Freitag16fast}, where $\vlambda^{l}$ is allowed to vary during learning. With our notations, this approach amounts to first learning an initial parameter value with all the data ($\forall d, \lambda^{l}(d) = \lambda^{s}(d)$), then to continue training with only batches from the test domain $d_t$ ($\lambda^{l}(d) = \indic{d = d_t}$) with $\indic{A}$ the indicator function for predicate $A$. Note that this strategy is potentially suboptimal, as some out-of-domain samples may contribute to the final performance due to eg.\ domain overlap. Optimizing the learning distribution in multi-domain settings is even more challenging, as the learner has to best take advantage of potential domains overlaps, and also of the fact that some domains might be easier to learn than others.\fyTodo{How to measure this?} \revision{Answer: The graphic of \system{Mixed-0} shows that with the same amount of data(batches) the convergences of the domains vary significantly}

@@

\section{Multi-Domain Automated Curriculum } \label{sec:mdac}
\subsection{Basic principles}
Assuming training data from $n_d$ domains $d_1 \dots d_{n_d}$, we denote the size of the training corpus from domain $d$ as  $N^{s}_d$, and $N^{s} = \sum_d N^{s}_d$ is the total number of training samples. We use $\widehat{\mathcal{D}^l_d}$ and $\widehat{\mathcal{D}^t _d}$ to denote the empirical train and test distributions for domain $d$, and $\widehat{\mathcal{D}^{u}}(x;\lambda^{u}) = \sum_{d} \lambda^{u}(d) \widehat{\mathcal{D}^{u}_d}(x)$ for $u\in\{l,t\}$. In our setting,  $\vlambda^t$, and hence $\widehat{\mathcal{D}^t}(x;\vlambda^t)$ are fixed and predefined, approximated with an equivalent number of development corpora.

Our method MDAC constructs an automatically adapting distribution $\lambda^{l}$ that optimizes the data selection policy along with the training of the NMT model. We parameterize $\lambda^{l}$ by a differentiable function $\lambda^l(\psi)$. We divide the training into many short sessions; in each session $t$, the model is trained with a fixed data distribution $\lambda^{l}(\psi_t)$. After one learning session, we update the data distribution via REINFORCE algorithm \citep{Williams92simple}.

\begin{align*}
\psi_{t+1} &= \psi_t + \mathbf{lr}_{data} * \displaystyle{\mathop{\sum}_{d=1}^K} R(d) * \frac{\partial \lambda^l(d;\psi_t)}{\partial \psi} \\
\end{align*}
\begingroup
\allowdisplaybreaks
With the following reward computation:
\begin{align*}
  R(d) = J^t(\theta_{t+k},\lambda^t) - J^t(\theta_t,\lambda^t)
\end{align*}
Where 
\begin{equation}
\begin{array}{rcl}
\theta_{t+i} &=& \theta_{t+i-1} - \mathbf{lr}_{nmt} \frac{1}{N}\displaystyle{\mathop{\sum}_{j=1}^N}\frac{\partial l(\theta_{t+i-1}, x^i_j,y^i_j)}{\partial \theta} \\ \nonumber
x^i_j, y^i_j &\sim& \widehat{D^l_d}(x) \\
J^t(\theta,\lambda_t) &=& \displaystyle{\mathop{\sum}_{d=1}^K}\lambda^t(d)\displaystyle{\mathop{\sum}_{x^t_d,y^t_d \in \widehat{D^t_d}}} l(\theta,x^t_d,y^t_d)
\end{array}
\end{equation}
\endgroup
$\mathbf{lr}_{data}$ is the learning rate of the parameter of the sampler. $\mathbf{lr}_{nmt}$ is the learning rate of the parameter of the NMT system. $J^t(\theta,\lambda_t)$ is the weighted loss computed over $K$ dev-sets corresponding to $K$ target domains.

To compute the reward of using data from domain $d$ to train the model, $R(d)$, we simulate k training steps from the current checkpoint by training the model with k batches sampled from $D^l(d)$ and compute the gain of the weighted dev-loss. The reward of our method is inspired by the target prediction gain in \cite{Graves17automated}. However, \cite{Graves17automated} used accumulated gains from the past as reward while we use the "future" gain as reward.

The work of \cite{Wang20balancing} based on the Bi-level Optimization framework, which aims to find an optimal fixed distribution $\lambda^{l}$ that will result in the best NMT model regarding a given target dev set at the end of the training, derives also a similar form of update for $\psi$. However, their reward is the cosine similarity between the gradient computed with the training data and the gradient computed with the dev set.

\subsection{MDAC for (multi) domain adaptation}
This setting is pretty general and can in principle accommodate the variety of situations mentioned  above, and many more: basic domain adaption, multi-domain adaptation with various target distributions, possibly including domains unseen in training. In our experiments we would like to better assess the actual potential of MDAC is these settings and seek to experimentally answer the following questions:
\begin{itemize}
\item is MDAC a viable alternative to conventional fine-tuning? In particular does it enable to better take advantage of relevant data from other domains?
\item is MDAC also a viable option in multi-domain adaptation scenarios?
\item does MDAC also enable to perform unsupervised (multi-) domain adaptation? \fyDone{TBContinued}
\end{itemize}

@@

\section{Experimental settings} \label{sec:exp}
\subsection{Data and metrics \label{ssec:corpora}}
We experiment with translation from English into French and use texts initially originating from 6~domains, corresponding to the following data sources: the UFAL Medical corpus V1.0 (\domain{med})\footnote{\url{https://ufal.mff.cuni.cz/ufal_medical_corpus}. We only use the in-domain (medical) subcorpora: PATR, EMEA, CESTA, ECDC.}; the European Central Bank corpus (\domain{bank}) \cite{Tiedemann12parallel}; the JRC-Acquis Communautaire corpus (\domain{law}) \cite{Steinberger06acquis}; documentations for KDE, Ubuntu, GNOME and PHP from the Opus collection \cite{Tiedemann09news}, merged in a \domain{it}-domain; TedTalks (\domain{talk}) \cite{Cettolo12wit}, and the Koran (\domain{rel}). Complementary experiments also use v12 of the News Commentary corpus (\domain{news}). Most corpora are available from the Opus website.\footnote{\url{http://opus.nlpl.eu}} These corpora were deduplicated and tokenized with in-house tools; statistics are in Table~\ref{tab:Corpora}. To reduce the number of types and build open-vocabulary systems, we use Byte-Pair Encoding \cite{Sennrich16BPE} with 30,000 merge operations on a corpus containing all sentences in both languages.\fyDone{Add \# number of tokens, also specificity ?}%

@@

@@

@@

\begin{itemize}
\itemsep0em 
\item Generic models trained with various predefined mixtures of the training data taking the form:
\begin{align} \label{mixture:trn}
\lambda_{\alpha}(d) = \frac{q_d^{\alpha}}{\displaystyle{\mathop{\sum}_{d=1}^{n_d}q_d^{\alpha}}} &&
q_d = \frac{\mid N^{s}_d \mid}{\displaystyle{N^{s}}} % \mathop{\sum}_{i=1}^K\mid D_i \mid}}
\end{align} \setlength{\abovedisplayskip}{2pt}
\setlength{\belowdisplayskip}{2pt} 
with $\alpha \in [0,0.25,0.5,0.75,1.0]$. These systems are denoted \system{Mixed-$\alpha$} below. \system{Mixed-$0$} uses a uniform domain distribution, \system{Mixed-$1.0$} simple uses the observed domain distribution.
\item fine-tuned models \cite{Luong15stanford,Freitag16fast}, based on the \system{Mixed-$1.0$} system, further trained on each domain for at most 50~000 iterations, with early stopping when the dev BLEU stops increasing. The full fine-tuning (\system{FT-Full}) procedure may update all the parameters of the initial generic model, resulting in six systems each adapted for one domain, with no parameter sharing across domains.
\item two multi-domain versions of the approach of \newcite{Bapna19simple}, denoted \system{FT-Res} and \system{MDL-Res}, where a domain-specific adaptation module is added to all the Transformer layers; within each layer, residual connections enable to short-cut this adapter. The former variant corresponds to the original proposal of \citet{Bapna19simple} (see also \cite{Sharaf20metalearning}). It fine-tunes the adapter modules of a \system{Mixed-$1.0$} system independently for each domain, keeping all the other parameters frozen. The latter uses the same architecture, but a different training procedure and learns all parameters jointly from scratch with the prefixed mixtures of training data \ref{mixture:trn}.\fyDone{Keep this ?}

\item Our comparison of multi-domain systems includes baselines with fixed data mixtures corresponding to $\vlambda_0$ and $\vlambda_1$, and our reimplementations of recent proposals from the literature: Curriculum Learning of \cite{Zhang19curriculum} and Differential Data Selection of \cite{Wang20balancing}.
\end{itemize}

@@

\subsection{MDAC systems} \label{ssec:dds-sys}
The behavior of MDAC only depends on (a) the initial domain distribution at the start of training $\vlambda^{l}_{t=0}$, and (b) the targeted (dev/test) distribution $\vlambda^{t}$. We will thus report these systems as \system{MDAC($\vlambda^{l}_{t=0}$, $\vlambda^{t}$)}. We also compare our method with the method DDS under the same conditions.

In our work, we simply parameterize the distribution $\lambda^l$ as follows.
\begin{equation}
\lambda
^l(d;\vpsi) = \frac{\vpsi[d]^\beta}{\sum_i \vpsi[i]^\beta} \nonumber
\end{equation}
We choose $\beta=2$ for our experiments.

We provide also in this section our choices of the hyper-parameters of our algorithm including the number of simulation training steps $k=10$, and the update rate $\mathbf{lr}_{data}=0.1$. We split the training into many short sessions that last 2000 training steps each.
\subsection{CL and DDS's re-implementation}
We re-implement DDS in Tensorflow without any change in the choices of parameterization and hyper-parameters compared to original code provided by \citet{Wang20balancing} \footnote{\url{https://github.com/cindyxinyiwang/multiDDS}}.

We re-implement the work of \citet{Zhang19curriculum} by following exactly the description of the algorithm. For each experiment of adapting the NMT system to one domain, we combine the training data of other domains into one corpus, then compute the cross-entropy difference score of each sentence of this corpus (we only compute the score for the source side). We sort and split the corpus into 9 shards and execute curriculum learning with 10 shards in which the first shard is the in-domain corpus.
\subsection{Experimental tasks}
We evaluate the efficacy of our methods compared to several aforementioned baselines via 3 tasks including domain adaptation; multi-domain adaptation and unsupervised domain adaptation.

In the domain adaptation task, given the data of 6 domains including \domain{med}, \domain{bank}, \domain{law}, \domain{it}, \domain{talk}, \domain{rel}, we aim to build expert NMT models for each domain respectively.

In the Multi-domain adaptation task, given the data of 6 domains, we aim to build an NMT model which gives the best performance regarding a predefined test set compromising equivalently 6 domains.

In the unsupervised domain adaptation task, we design 2 experiments. In the first experiment, given the data of 6 domains, we aim to build an NMT model which is robust to an unseen domain, which is News in this case. In the second experiment, we cluster the training data of 6 domains into 30 clusters using the KNN algorithm in the same way as in the work of \citet{Tars18multidomain}, then adapt these clusters to one of 6 domains given the corresponding in-domain dev set. We compare MDAC to the contrast DDS over 6 domains.

@@

\subsection{Domain Adaptation}
In this setting, we aim to build an expert NMT model in only one domain. This setting corresponds to Dirac distribution $\lambda^t$, in which the probability of domains is 0 except for one domain.

We initialize MDAC and DDS by 2 different distribution $\lambda_0$ and $\lambda_1$. According to table \ref{tab:da}, MDAC achieves the best performance with $\lambda_0$ initial where all domains have the same probability to be chosen. The same conclusion is also true for DDS.
